---
title: "CytoFlag"
author:
- name: Tim Mocking
  affiliation: AmsterdamUMC Cancer Center
- name: Yejin Park
  affiliation: AmsterdamUMC Cancer Center
- name: Costa Bachas
  affiliation: AmsterdamUMC Cancer Center
# package: CytoFlag
output:
  # BiocStyle::pdf_document
  BiocStyle::html_document
abstract: |
  This vignette describes a workflow for analyzing cytometry data using CytoFlag.
vignette: |
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Installation

```{r}
# Example for when Bioconductor installation is available
```

# Setting up a CytoFlag workflow

The CytoFlag workflow takes file paths of .fcs files as input. 

We provide the generateDemo function which generates a folder containing a small
set of FCS files which is used in this tutorial.

## Loading the demo data

```{r}
# Create a folder "demo_data" containing 100 small FCS files for demonstration
generateDemo(dir = "demo_data", nFiles = 100, nCells = 1000, nChannels = 8)

# List the generated files
files <- list.files("demo_data", full.names = TRUE)
head(files)
```

## Initializing a CytoFlag object

The first step is always to initialize a CytoFlag object. This object stores all 
relevant information, from file paths to generated features and outputs.

```{r}
CF <- CytoFlag()
```

## Adding data to a CytoFlag object

### Adding cytometry data

If you want to use CytoFlag for outlier detection, you add all the data as 
"test data". This means that all models are trained and outliers are detected
on the full dataset.

```{r}
CF <- CytoFlag()
CF <- addTestdata(CF, files) 
```

If you want to use CytoFlag for novelty, you have to define reference data as
well. In this case, models will be trained on the reference data, before identifying
novelties in the test data.

```{r}
CF <- CytoFlag()
CF <- addReferencedata(CF, files[1:50]) 
CF <- addTestdata(CF, files[51:100]) 
```

### Adding labels

If you want to visualize different known variables in CytoFlag visualizations,
you have to add these to the CytoFlag object as follows:

```{r}
# Generate some random batches
refLabels <- sample(c("batch1", "batch2"), 50, replace = TRUE)
testLabels <- sample(c("batch1", "batch2"), 50, replace = TRUE)
head(refLabels)
```

```{r}
# Add them to the CytoFlag object
CF <- addReferencelabels(CF, refLabels)
CF <- addTestlabels(CF, testLabels)
```

### Custom pre-processing

By default, CytoFlag uses the same pre-processing function for all loaded 
flowframes. 

You can modify this by supplying your own pre-processing function, for example
by only performing a MinMax normalization.

```{r}
CF <- CytoFlag()

# Define a custom function which also scales expression values from 0 to 1
MinMax <- function(ff){
  spill <- ff@description$SPILL
  ff <- flowCore::compensate(ff, spill)
  ff <- flowCore::transform(ff, flowCore::transformList(colnames(spill), 
                                            flowCore::arcsinhTransform(a = 0, 
                                                                       b = 1/150, 
                                                                       c = 0)))
  ff@exprs[, colnames(spill)] <- apply(ff@exprs[, colnames(spill)], 2, function(x){
    return((x - quantile(x, 0.01)) / (quantile(x, 0.99) - quantile(x, 0.01)))
  })
  return(ff)
}

# Replace the default pre-processing
CF$preprocessFunction <- MinMax

# Read some data
CF <- addTestdata(CF, files[1], read = TRUE, aggSize = 1000)

# Plot a histogram of the first marker
hist(CF$data$test[[files[1]]][, 1], main = files[1], xlab = "Scaled expression")
```

## Feature generation

### Generating features for outlier detection

```{r}
CF <- CytoFlag() 
CF <- addTestdata(CF, files)
# The channels of the simulated data are letters of the alphabet
channels <- LETTERS[1:8]
CF <- generateFeatures(CF, channels = channels, featMethod = "quantiles")
head(CF$features$test$quantiles)
```

### Generating features for novelty detection

In this example, we use the earth mover's distance (EMD). Because these distances
are calculated against an aggregated set of cells.

If reference data is present, this is automatically used for aggregation.

```{r, eval=FALSE}
CF <- CytoFlag()
CF <- addReferencedata(CF, files[51:100]) 
CF <- addTestdata(CF, files[1:50]) 
CF <- generateFeatures(CF, channels = channels, featMethod = "EMD")
head(CF$features$test$EMD)
```

## Flagging anomalies

### Flagging anomalies using outlier detection

```{r}
CF <- CytoFlag() 
CF <- addTestdata(CF, files)
# The channels of the simulated data are letters of the alphabet
channels <- LETTERS[1:8]
CF <- generateFeatures(CF, channels = channels, featMethod = "quantiles")
CF <- Flag(CF, featMethod = "quantiles", flagStrat = "outlier")
head(CF$outliers$quantiles)
```

### Flagging anomalies using novelty detection

TO-DO

## Visualizations

```{r}
CF <- CytoFlag() 
CF <- addReferencedata(CF, files[0:50])
CF <- addTestdata(CF, files[51:100])
# The channels of the simulated data are letters of the alphabet
channels <- LETTERS[1:8]
CF <- generateFeatures(CF, channels = channels, featMethod = "quantiles",
                       quantileDist = 0.2)
```

```{r}
plotHeatmap(CF, featMethod = "quantiles", plotRef = TRUE)
```
```{r}
# plotPCA(CF, featMethod = "quantiles", plotRef = TRUE)
```

### Visualizing marker distributions

### Visualizing features in 2D space (PCA)

### Visualizing features using heatmaps

### Visualizing marker distributions of anomalous samples


